Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2024-07-02 manuscript/manuscript.Rmd
Additional examples.

Invariant power transformations are intended to yield improved transformations to normality
in the presence of large shifts in location, scales far from unity, or both.
We previously demonstrated this property using simulations, and will now demonstrate
it using examples from real datasets.

### Age in a ...

In health-related datasets patient age is a common characteristic.

Show invariance to units (days vs years)

### Body mass of penguins
modeldata::penguins$body_mass_g

@@

```{r top-gear, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for the fuel consumption data in the Top Gear dataset. ",
  "The dataset contains three strong outliers for highly fuel-efficient vehicles.",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The empirical central normality test indicates that the original data do not deviate significantly from central normality ($p=0.76$).",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.55$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations. ",
  "For these, the empirical central normality test yields $p=0.03$, and $p=0.86$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

.plot_top_gear(
  plot_theme = plot_theme
)
```

@@

.plot_ischemic_stroke(
  plot_theme = plot_theme
)
```
The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
Angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Figure \ref{fig:ischemic-stroke} show that non-robust transformation fail to produce a centrally normal distribution (empirical central normality test $p = 0.01$ and $p=0.02$ for conventional and invariant transformations, respectively). Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

>>>>

Invariant power transformations are intended to yield improved transformations to normality
in the presence of large shifts in location, scales far from unity, or both.
Above, this characteristic was assessed using simulations. Below, this characteristic
will be evaluated using examples from real datasets.

### Age of patients with lung cancer

A common variable in health-related datasets is age.
Here we use data on 228 patients with lung cancer that was collected and published by Loprinzi et al. [@Loprinzi1994-cd].
The age in the cohort was $62.4 \pm 9.1$ (mean ± standard deviation) years.
Applying conventional and invariant Yeo-Johnson transformations to patient age yielded the following results, see Figure \ref{fig:lung-cancer-age}:
no transformation (residual sum $\sum r_i = 16.5$);
conventional transformation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$);
shift- and scale invariant transformation ($\lambda = 0.9$, $\sum r_i = 8.8$, $\mu_{YJ} = 1.2$, $\sigma_{YJ} = 1.1$);
and robust shift- and scale-invariant transformation ($\lambda = 0.8$, $\sum r_i = 10.6$, $\mu_{YJ} = 1.2$, $\sigma_{YJ} = 1.0$).

```{r lung-cancer-age, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for age of patients with lung cancer, expressed in years. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations."
)

out <- .plot_lung_cancer_age(plot_theme = plot_theme)
out$plot
```

Shift- and scale-invariant transformation leads to a lower overall residual sum, indicating a better transformation.
Moreover, conventional transformations inflated the mean $\mu_{YJ}$ and standard deviation $\sigma_{YJ}$ of the age variable after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.78$).

### Penguin body mass

Gorman, Williams and Fraser recorded body mass (in grams) of 342 penguins of three different species [@Gorman2014-eo].
The body mass was $(4.2 \pm 0.8) \cdot 10^3$ (mean ± standard deviation) grams.
Applying conventional and invariant Yeo-Johnson transformations to patient age yielded the following results, see Figure \ref{fig:penguin-body-mass}:
no transformation (residual sum $\sum r_i = 48.0$);
conventional transformation ($\lambda = 0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.102$, $\sigma_{YJ} = 4 \cdot 10^{-3}$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.101$, $\sigma_{YJ} = 4 \cdot 10^{-3}$);
shift- and scale invariant transformation ($\lambda = 0.5$, $\sum r_i = 26.8$, $\mu_{YJ} = 0.9$, $\sigma_{YJ} = 0.9$);
and robust shift- and scale-invariant transformation ($\lambda = 0.4$, $\sum r_i = 23.1$, $\mu_{YJ} = 0.8$, $\sigma_{YJ} = 0.8$).

```{r penguin-body-mass, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for penguin body mass, measured in grams. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations."
)

out <- .plot_penguin_body_mass(plot_theme = plot_theme)
out$plot
```

@@

```{r top-gear, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Quantile-quantile plots for fuel consumption data in the Top Gear dataset. ",
  "The dataset contains three strong outliers for highly fuel-efficient vehicles.",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The empirical central normality test indicates that the original data do not deviate significantly from central normality ($p=0.76$).",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation \\cite{Raymaekers2021-kq}. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.55$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust invariant transformations. ",
  "For these, the empirical central normality test yields $p=0.03$, and $p=0.86$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

out <- .plot_top_gear(
  plot_theme = plot_theme
)

out$plot
```

@@

out <- .plot_ischemic_stroke(
  plot_theme = plot_theme
)

out$plot
```
The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
Angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Figure \ref{fig:ischemic-stroke} show that non-robust transformation fail to produce a centrally normal distribution (empirical central normality test $p = 0.01$ and $p=0.02$ for conventional and invariant transformations, respectively). Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).
